### Northern Ireland: validation case {#sec-results-ni}

Northern Ireland serves as the validation case: the Troubles are well-documented, independent conflict-death tallies exist, and HMD data quality is high. The model's estimates can be directly compared against these tallies to assess calibration.

```{r}
#| label: load-data
#| include: false

# Load HMD data
hmd <- read_csv("data/processed/hmd_counts.csv", show_col_types = FALSE)
ni <- hmd |> filter(country_code == "GBR_NIR")

# Prepare NI male data for modelling
ni_male <- ni |>
  filter(sex == "male") |>
  filter(!is.na(deaths), !is.na(exposure)) |>
  prepare_model_data(sex_filter = "male", age_range = c(AGE_MIN, AGE_MAX), k = OPTIMAL_DECAY_K)

# Fit the full model (model 5 from original analysis)
mod_full <- fit_troubles_model(ni_male)

# Predictions with Troubles
ni_male_preds <- ni_male |>
  modelr::add_predictions(mod_full, var = "pred_full")

# Counterfactual predictions (no Troubles)
ni_male_counter <- ni_male |>
  mutate(trbls_kpc = 0) |>
  broom::augment(mod_full, newdata = _) |>
  select(year, age, pred_counter = .fitted)

# Combine
ni_decomp <- ni_male_preds |>
  left_join(ni_male_counter, by = c("year", "age")) |>
  mutate(impulse_decay = pred_full - pred_counter)
```

#### Raw death counts by sex

Before any modelling, the conflict signature is visible in raw data. @fig-ni-raw-counts shows total deaths for males and females aged `r AGE_MIN`--`r AGE_MAX` in Northern Ireland. The post-1972 divergence — a sustained uptick in male deaths with no corresponding female increase — is the pattern the impulse-decay model formalises.

```{r}
#| label: fig-ni-raw-counts
#| fig-cap: "Total deaths by sex for ages 15--45 in Northern Ireland, 1922--2023. The vertical dashed line marks 1972 (onset of the Troubles). Male deaths rise sharply after 1972 and remain elevated for decades; female deaths show no corresponding increase."
#| fig-width: 8
#| fig-height: 5

ni |>
  filter(!is.na(year), !is.na(deaths),
         age >= AGE_MIN, age <= AGE_MAX) |>
  group_by(year, sex) |>
  summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
  mutate(sex = str_to_title(sex)) |>
  ggplot(aes(x = year, y = deaths, colour = sex, shape = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_line(linewidth = 0.4, alpha = 0.5) +
  geom_vline(xintercept = 1972, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Year", y = "Total deaths",
       colour = "Sex", shape = "Sex",
       subtitle = paste0("Ages ", AGE_MIN, "--", AGE_MAX)) +
  theme_minimal(base_size = 12)
```

#### Visual decomposition

@fig-ni-decomposition presents the four-panel decomposition for males aged `r AGE_MIN`--`r AGE_MAX` in Northern Ireland, 1922--2023.

```{r}
#| label: fig-ni-decomposition
#| fig-cap: "Decomposition of the mortality surface for males aged 15--45 in Northern Ireland. (A) Observed log₁₀ mortality rates. (B) Full model: smooth baseline plus impulse-decay component. (C) Counterfactual: smooth baseline with impulse-decay turned off. (D) Impulse-decay component alone (B minus C), representing estimated conflict-attributable excess mortality."
#| fig-width: 10
#| fig-height: 10

lmr_range <- range(c(ni_decomp$lmr, ni_decomp$pred_full, ni_decomp$pred_counter), na.rm = TRUE)

panel_a <- ggplot(ni_decomp, aes(x = year, y = age, fill = lmr)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "A: Observed", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_b <- ggplot(ni_decomp, aes(x = year, y = age, fill = pred_full)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "B: Full model", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_c <- ggplot(ni_decomp, aes(x = year, y = age, fill = pred_counter)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "C: Counterfactual", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_d <- ggplot(ni_decomp |> filter(year >= 1960),
                  aes(x = year, y = age, fill = impulse_decay)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "firebrick",
                       midpoint = 0, name = "Δ log₁₀(m)") +
  labs(title = "D: Impulse-decay component", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

patchwork <- (panel_a + panel_b) / (panel_c + panel_d)
patchwork
```

Panel A shows the raw data, including the visible band of elevated mortality in young adult males after 1972. Panel B shows the model's stylised representation — the reader can assess fit by comparing A and B. Panel C shows the counterfactual: what the surface would have looked like had the Troubles not occurred. Panel D isolates the impulse-decay component itself — the conflict's estimated demographic footprint.

#### Parameter estimates

```{r}
#| label: tbl-ni-params
#| tbl-cap: "Estimated impulse-decay parameters for Northern Ireland"

# Optimal decay rate (pre-computed via AIC optimisation)
optimal <- find_optimal_decay(ni, criterion = "AIC")
k_opt <- optimal$minimum
halflife <- log(0.5) / log(1 - k_opt)
time_to_5pct <- log(0.05) / log(1 - k_opt)

# Extract Troubles coefficients by age
coefs <- coef(mod_full)
trbls_coefs <- coefs[grep("trbls_kpc", names(coefs))]
# First is the base (age 15), rest are interactions
base_effect <- trbls_coefs[1]
interaction_effects <- trbls_coefs[-1]
age_effects <- c(base_effect, base_effect + interaction_effects)
names(age_effects) <- AGE_MIN:AGE_MAX
peak_age <- as.integer(names(which.max(abs(age_effects))))
peak_effect <- max(abs(age_effects))

tibble(
  Parameter = c("Onset year", "Decay rate (k)", "Half-life (years)",
                "Peak age", "Peak impulse (log₁₀ units)",
                "Time to 5% intensity (years)"),
  Estimate = c(
    as.character(TROUBLES_START_YEAR),
    sprintf("%.4f", k_opt),
    sprintf("%.1f", halflife),
    as.character(peak_age),
    sprintf("%.3f", peak_effect),
    sprintf("%.0f (~%d)", time_to_5pct, round(TROUBLES_START_YEAR + time_to_5pct))
  )
) |>
  knitr::kable()
```

#### Cumulative excess deaths

@fig-ni-cumulative shows the cumulative estimated excess deaths over time, derived by summing the impulse-decay component (panel D) across ages, weighted by population exposure.

```{r}
#| label: fig-ni-cumulative
#| fig-cap: "Cumulative estimated excess male deaths (ages 15--45) attributable to the Troubles. Dashed line: approximate cumulative directly-attributed conflict deaths from Smyth (1998)."

excess <- estimate_excess_deaths(ni, mod_full, ni_male)

excess_by_year <- excess |>
  filter(year >= 1970) |>
  group_by(year) |>
  summarise(annual_excess = sum(excess, na.rm = TRUE), .groups = "drop") |>
  mutate(cumulative = cumsum(annual_excess))

total_excess <- round(max(excess_by_year$cumulative))

# Smyth (1998) tally: ~3,598 total; roughly scaled to males 15-45
smyth_total <- 3598

ggplot(excess_by_year, aes(x = year, y = cumulative)) +
  geom_line(linewidth = 1, colour = "firebrick") +
  geom_hline(yintercept = smyth_total, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 2010, y = smyth_total + 100,
           label = paste0("Smyth (1998): ", format(smyth_total, big.mark = ",")),
           size = 3.5, colour = "grey40") +
  annotate("text", x = 2010, y = total_excess - 200,
           label = paste0("Model estimate: ", format(total_excess, big.mark = ",")),
           size = 3.5, colour = "firebrick") +
  labs(x = "Year", y = "Cumulative excess deaths",
       subtitle = paste0("Males aged ", AGE_MIN, "-", AGE_MAX,
                         "; model captures ~",
                         round(100 * total_excess / smyth_total), "% of tallied deaths")) +
  theme_minimal(base_size = 12)
```

The comparison with the independent tally from @smyth1998 provides a direct calibration check. Our model, using only all-cause mortality data, is expected to capture approximately 75% of the directly-attributed deaths — the remainder reflecting mortality displacement effects, emigration as a competing risk, and the restriction to males aged `r AGE_MIN`--`r AGE_MAX` (see @sec-comparison).

#### Troubles effect by age and sex

As a specificity check, the same model is applied to the female mortality surface. @fig-ni-sex-comparison shows the impulse-decay coefficient by age for both sexes.

```{r}
#| label: fig-ni-sex-comparison
#| fig-cap: "Impulse-decay coefficient by age for males and females in Northern Ireland. The conflict signature is concentrated in young adult males, with a peak near age 18--20. The female surface shows negligible or no conflict effect, consistent with the sex-specific nature of sectarian violence."

# Male coefficients
male_effects <- tibble(
  age = AGE_MIN:AGE_MAX,
  effect = as.numeric(age_effects),
  sex = "Male"
)

# Female model
ni_female <- ni |>
  filter(sex == "female") |>
  filter(!is.na(deaths), !is.na(exposure)) |>
  prepare_model_data(sex_filter = "female", age_range = c(AGE_MIN, AGE_MAX), k = OPTIMAL_DECAY_K)

mod_female <- fit_troubles_model(ni_female)
coefs_f <- coef(mod_female)
trbls_coefs_f <- coefs_f[grep("trbls_kpc", names(coefs_f))]
base_f <- trbls_coefs_f[1]
inter_f <- trbls_coefs_f[-1]
age_effects_f <- c(base_f, base_f + inter_f)

female_effects <- tibble(
  age = AGE_MIN:AGE_MAX,
  effect = as.numeric(age_effects_f),
  sex = "Female"
)

bind_rows(male_effects, female_effects) |>
  ggplot(aes(x = age, y = effect, colour = sex, shape = sex)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 18, linetype = "dotted", alpha = 0.5) +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Age", y = "Impulse-decay coefficient",
       colour = "Sex", shape = "Sex") +
  theme_minimal(base_size = 12)
```

#### Residual surface

```{r}
#| label: fig-ni-residuals
#| fig-cap: "Residual surface (predicted minus observed) for the full model fitted to Northern Ireland males aged 15--45. The absence of structured residual patterns after 1972 indicates that the impulse-decay component adequately captures the conflict signature."

ni_decomp |>
  mutate(residual = pred_full - lmr) |>
  ggplot(aes(x = year, y = age, fill = residual)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0,
                       limits = c(-0.6, 0.6),
                       oob = scales::squish,
                       name = "Residual") +
  labs(x = "Year", y = "Age") +
  theme_minimal(base_size = 12) +
  coord_fixed(ratio = 1)
```

#### Model comparison

```{r}
#| label: tbl-ni-model-comparison
#| tbl-cap: "Model comparison for Northern Ireland males. The inclusion of the impulse-decay term (Model 5) substantially improves fit, with a ~19% reduction in root mean square error."

# Fit models without Troubles for comparison
mod_no_troubles <- lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
                        yr_since_imp + yr_since_imp2),
                      data = ni_male)

rms_no_troubles <- rms(residuals(mod_no_troubles))
rms_full <- rms(residuals(mod_full))

tibble(
  Model = c("Baseline (no Troubles term)", "Full (with impulse-decay)"),
  AIC = c(AIC(mod_no_troubles), AIC(mod_full)),
  BIC = c(BIC(mod_no_troubles), BIC(mod_full)),
  RMSE = c(rms_no_troubles, rms_full),
  `RMSE reduction` = c("—", sprintf("%.1f%%", 100 * (1 - rms_full / rms_no_troubles)))
) |>
  mutate(across(c(AIC, BIC), \(x) sprintf("%.0f", x)),
         RMSE = sprintf("%.4f", RMSE)) |>
  knitr::kable()
```

---

### East and West Germany {#sec-results-germany}

East and West Germany provide an instructive contrast. Both populations experienced the aftermath of WWII, but diverged sharply after partition: East Germany (DEUTE) under Soviet occupation and state socialism; West Germany (DEUTW) under Western integration and rapid economic growth. The HMD data cover 1956--2021 for both, beginning after the immediate post-war period. Previous work has constructed a "Synthetic Germany" series by weighting East and West German life expectancy to approximate a continuous pre-reunification national trajectory [@mintonhiam2022]; here we examine each component separately to characterise their distinct mortality dynamics.

Because the German data begin in 1956 — after the NI model's pre-1938 and early phases — a simplified baseline is used: a quadratic age-year trend with the same impulse-decay overlay. Rather than fixing the onset at 1972, we scan over candidate onset years for each population and select the onset that minimises AIC, allowing the data to identify the most salient disruption.

```{r}
#| label: germany-setup
#| include: false

# Helper: fit and decompose for a non-NI country with simplified baseline
fit_country_case <- function(country_data, k = OPTIMAL_DECAY_K,
                             onset = TROUBLES_START_YEAR,
                             age_range = c(AGE_MIN, AGE_MAX)) {
  fit_one_sex <- function(dta, sex_val) {
    sex_data <- dta |>
      filter(sex == sex_val, !is.na(deaths), !is.na(exposure),
             age >= age_range[1], age <= age_range[2]) |>
      mutate(
        lmr = calc_lmr(deaths, exposure),
        yr_centred = year - median(year),
        impulse = ifelse(year < onset, 0, (1 - k) ^ (year - onset))
      )

    mod0 <- lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2)),
               data = sex_data)
    mod1 <- lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2) + impulse),
               data = sex_data)

    preds <- sex_data |>
      modelr::add_predictions(mod1, var = "pred_full")
    counter <- sex_data |>
      mutate(impulse = 0) |>
      broom::augment(mod1, newdata = _) |>
      select(year, age, pred_counter = .fitted)

    decomp <- preds |>
      left_join(counter, by = c("year", "age")) |>
      mutate(impulse_decay = pred_full - pred_counter,
             residual = pred_full - lmr)

    # Extract impulse coefficients by age
    coefs <- coef(mod1)
    imp_coefs <- coefs[grep("impulse", names(coefs))]
    base_imp <- imp_coefs[1]
    inter_imp <- if (length(imp_coefs) > 1) imp_coefs[-1] else numeric(0)
    age_effects <- c(base_imp, base_imp + inter_imp)
    names(age_effects) <- age_range[1]:age_range[2]

    list(mod0 = mod0, mod1 = mod1, decomp = decomp,
         age_effects = age_effects, sex = sex_val)
  }

  male <- fit_one_sex(country_data, "male")
  female <- fit_one_sex(country_data, "female")

  list(male = male, female = female, data = country_data,
       onset = onset, k = k)
}

# Scan over onset years to find optimal for a given country
find_optimal_onset <- function(country_data, k = OPTIMAL_DECAY_K,
                                onset_range = NULL,
                                age_range = c(AGE_MIN, AGE_MAX)) {
  male_data <- country_data |>
    filter(sex == "male", !is.na(deaths), !is.na(exposure),
           age >= age_range[1], age <= age_range[2]) |>
    mutate(lmr = calc_lmr(deaths, exposure),
           yr_centred = year - median(year))

  if (is.null(onset_range)) {
    yr_range <- range(male_data$year)
    # Need at least 5 years before and after onset
    onset_range <- seq(yr_range[1] + 5, yr_range[2] - 5)
  }

  aic_by_onset <- map_dfr(onset_range, function(t0) {
    d <- male_data |>
      mutate(impulse = ifelse(year < t0, 0, (1 - k)^(year - t0)))
    mod <- tryCatch(
      lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2) + impulse), data = d),
      error = function(e) NULL)
    if (is.null(mod)) return(NULL)
    tibble(onset = t0, aic = AIC(mod))
  })

  aic_by_onset
}

# Fit East and West Germany with onset scanning
deute_data <- hmd |> filter(country_code == "DEUTE")
deutw_data <- hmd |> filter(country_code == "DEUTW")

deute_onset_scan <- find_optimal_onset(deute_data)
deutw_onset_scan <- find_optimal_onset(deutw_data)

deute_best_onset <- deute_onset_scan |> slice_min(aic) |> pull(onset)
deutw_best_onset <- deutw_onset_scan |> slice_min(aic) |> pull(onset)

deute_fit <- fit_country_case(deute_data, onset = deute_best_onset)
deutw_fit <- fit_country_case(deutw_data, onset = deutw_best_onset)
```

#### Onset year scan

@fig-germany-onset-scan shows the AIC profile over candidate onset years for East and West Germany. The data-driven onset identifies the most salient mortality disruption for each population.

```{r}
#| label: fig-germany-onset-scan
#| fig-cap: "AIC profile over candidate onset years for East and West Germany. East Germany shows a clear minimum near 1990 (reunification), while West Germany's profile identifies a different optimal onset."
#| fig-width: 8
#| fig-height: 5

bind_rows(
  deute_onset_scan |> mutate(country = "East Germany"),
  deutw_onset_scan |> mutate(country = "West Germany")
) |>
  group_by(country) |>
  mutate(delta_aic = aic - min(aic)) |>
  ungroup() |>
  ggplot(aes(x = onset, y = delta_aic, colour = country)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 1990, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 1990, y = Inf, label = "Reunification", vjust = 1.5,
           size = 3, colour = "grey40") +
  geom_vline(xintercept = 1972, linetype = "dashed", alpha = 0.3) +
  annotate("text", x = 1972, y = Inf, label = "1972 (NI onset)", vjust = 3,
           size = 2.5, colour = "grey60") +
  scale_colour_manual(values = c("East Germany" = "firebrick",
                                  "West Germany" = "steelblue")) +
  labs(x = "Candidate onset year", y = "ΔAIC (relative to best)",
       colour = NULL) +
  coord_cartesian(ylim = c(0, 500)) +
  theme_minimal(base_size = 12)
```

For East Germany, the optimal onset is `r deute_best_onset`, consistent with the shock of reunification. For West Germany, the optimal onset is `r deutw_best_onset`. The remaining analysis uses these data-driven onsets.

#### East Germany (DEUTE)

```{r}
#| label: fig-deute-raw-counts
#| fig-cap: "Total deaths by sex for ages 15--45 in East Germany, 1956--2021. Vertical dashed line at 1972."
#| fig-width: 8
#| fig-height: 5

deute_data |>
  filter(!is.na(year), !is.na(deaths),
         age >= AGE_MIN, age <= AGE_MAX) |>
  group_by(year, sex) |>
  summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
  mutate(sex = str_to_title(sex)) |>
  ggplot(aes(x = year, y = deaths, colour = sex, shape = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_line(linewidth = 0.4, alpha = 0.5) +
  geom_vline(xintercept = 1972, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Year", y = "Total deaths", colour = "Sex", shape = "Sex",
       subtitle = paste0("East Germany, ages ", AGE_MIN, "--", AGE_MAX)) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-deute-decomposition
#| fig-cap: "Decomposition of the mortality surface for males aged 15--45 in East Germany. (A) Observed. (B) Full model. (C) Counterfactual. (D) Impulse-decay component."
#| fig-width: 10
#| fig-height: 10

deute_decomp <- deute_fit$male$decomp
deute_lmr_range <- range(c(deute_decomp$lmr, deute_decomp$pred_full,
                            deute_decomp$pred_counter), na.rm = TRUE)

p_a <- ggplot(deute_decomp, aes(x = year, y = age, fill = lmr)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deute_lmr_range, name = "log₁₀(m)") +
  labs(title = "A: Observed", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_b <- ggplot(deute_decomp, aes(x = year, y = age, fill = pred_full)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deute_lmr_range, name = "log₁₀(m)") +
  labs(title = "B: Full model", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_c <- ggplot(deute_decomp, aes(x = year, y = age, fill = pred_counter)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deute_lmr_range, name = "log₁₀(m)") +
  labs(title = "C: Counterfactual", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_d <- ggplot(deute_decomp |> filter(year >= 1960),
              aes(x = year, y = age, fill = impulse_decay)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "firebrick",
                       midpoint = 0, name = "Δ log₁₀(m)") +
  labs(title = "D: Impulse-decay component", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

(p_a + p_b) / (p_c + p_d)
```

```{r}
#| label: fig-deute-sex-comparison
#| fig-cap: "Impulse-decay coefficient by age for males and females in East Germany."
#| fig-width: 8
#| fig-height: 5

bind_rows(
  tibble(age = AGE_MIN:AGE_MAX,
         effect = as.numeric(deute_fit$male$age_effects), sex = "Male"),
  tibble(age = AGE_MIN:AGE_MAX,
         effect = as.numeric(deute_fit$female$age_effects), sex = "Female")
) |>
  ggplot(aes(x = age, y = effect, colour = sex, shape = sex)) +
  geom_point(size = 2) + geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Age", y = "Impulse-decay coefficient", colour = "Sex", shape = "Sex",
       subtitle = "East Germany") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-deute-residuals
#| fig-cap: "Residual surface for the full model fitted to East Germany males aged 15--45."
#| fig-width: 8
#| fig-height: 5

deute_decomp |>
  ggplot(aes(x = year, y = age, fill = residual)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-0.6, 0.6),
                       oob = scales::squish, name = "Residual") +
  labs(x = "Year", y = "Age") +
  theme_minimal(base_size = 12) + coord_fixed(ratio = 1)
```

```{r}
#| label: tbl-deute-params
#| tbl-cap: "Model comparison and parameter estimates for East Germany males."

deute_rms0 <- rms(residuals(deute_fit$male$mod0))
deute_rms1 <- rms(residuals(deute_fit$male$mod1))
deute_peak_age <- as.integer(names(which.max(abs(deute_fit$male$age_effects))))
deute_peak_effect <- max(abs(deute_fit$male$age_effects))
deute_male_mean <- mean(abs(deute_fit$male$age_effects))
deute_female_mean <- mean(abs(deute_fit$female$age_effects))

tibble(
  Measure = c(
    "Onset year (data-driven)",
    "AIC (baseline)", "AIC (full model)", "AIC improvement",
    "RMSE (baseline)", "RMSE (full model)", "RMSE reduction",
    "Peak age", "Peak impulse (log₁₀ units)",
    "Mean male impulse", "Mean female impulse", "M/F ratio"
  ),
  Value = c(
    as.character(deute_best_onset),
    sprintf("%.0f", AIC(deute_fit$male$mod0)),
    sprintf("%.0f", AIC(deute_fit$male$mod1)),
    sprintf("%.0f", AIC(deute_fit$male$mod0) - AIC(deute_fit$male$mod1)),
    sprintf("%.4f", deute_rms0),
    sprintf("%.4f", deute_rms1),
    sprintf("%.1f%%", 100 * (1 - deute_rms1 / deute_rms0)),
    as.character(deute_peak_age),
    sprintf("%.3f", deute_peak_effect),
    sprintf("%.3f", deute_male_mean),
    sprintf("%.3f", deute_female_mean),
    sprintf("%.2f", deute_male_mean / deute_female_mean)
  )
) |>
  knitr::kable()
```

#### West Germany (DEUTW)

```{r}
#| label: fig-deutw-raw-counts
#| fig-cap: "Total deaths by sex for ages 15--45 in West Germany, 1956--2021. Vertical dashed line at 1972."
#| fig-width: 8
#| fig-height: 5

deutw_data |>
  filter(!is.na(year), !is.na(deaths),
         age >= AGE_MIN, age <= AGE_MAX) |>
  group_by(year, sex) |>
  summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
  mutate(sex = str_to_title(sex)) |>
  ggplot(aes(x = year, y = deaths, colour = sex, shape = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_line(linewidth = 0.4, alpha = 0.5) +
  geom_vline(xintercept = 1972, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Year", y = "Total deaths", colour = "Sex", shape = "Sex",
       subtitle = paste0("West Germany, ages ", AGE_MIN, "--", AGE_MAX)) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-deutw-decomposition
#| fig-cap: "Decomposition of the mortality surface for males aged 15--45 in West Germany. (A) Observed. (B) Full model. (C) Counterfactual. (D) Impulse-decay component."
#| fig-width: 10
#| fig-height: 10

deutw_decomp <- deutw_fit$male$decomp
deutw_lmr_range <- range(c(deutw_decomp$lmr, deutw_decomp$pred_full,
                            deutw_decomp$pred_counter), na.rm = TRUE)

p_a <- ggplot(deutw_decomp, aes(x = year, y = age, fill = lmr)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deutw_lmr_range, name = "log₁₀(m)") +
  labs(title = "A: Observed", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_b <- ggplot(deutw_decomp, aes(x = year, y = age, fill = pred_full)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deutw_lmr_range, name = "log₁₀(m)") +
  labs(title = "B: Full model", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_c <- ggplot(deutw_decomp, aes(x = year, y = age, fill = pred_counter)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = deutw_lmr_range, name = "log₁₀(m)") +
  labs(title = "C: Counterfactual", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

p_d <- ggplot(deutw_decomp |> filter(year >= 1960),
              aes(x = year, y = age, fill = impulse_decay)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "firebrick",
                       midpoint = 0, name = "Δ log₁₀(m)") +
  labs(title = "D: Impulse-decay component", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) + coord_fixed(ratio = 1)

(p_a + p_b) / (p_c + p_d)
```

```{r}
#| label: fig-deutw-sex-comparison
#| fig-cap: "Impulse-decay coefficient by age for males and females in West Germany."
#| fig-width: 8
#| fig-height: 5

bind_rows(
  tibble(age = AGE_MIN:AGE_MAX,
         effect = as.numeric(deutw_fit$male$age_effects), sex = "Male"),
  tibble(age = AGE_MIN:AGE_MAX,
         effect = as.numeric(deutw_fit$female$age_effects), sex = "Female")
) |>
  ggplot(aes(x = age, y = effect, colour = sex, shape = sex)) +
  geom_point(size = 2) + geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Age", y = "Impulse-decay coefficient", colour = "Sex", shape = "Sex",
       subtitle = "West Germany") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-deutw-residuals
#| fig-cap: "Residual surface for the full model fitted to West Germany males aged 15--45."
#| fig-width: 8
#| fig-height: 5

deutw_decomp |>
  ggplot(aes(x = year, y = age, fill = residual)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-0.6, 0.6),
                       oob = scales::squish, name = "Residual") +
  labs(x = "Year", y = "Age") +
  theme_minimal(base_size = 12) + coord_fixed(ratio = 1)
```

```{r}
#| label: tbl-deutw-params
#| tbl-cap: "Model comparison and parameter estimates for West Germany males."

deutw_rms0 <- rms(residuals(deutw_fit$male$mod0))
deutw_rms1 <- rms(residuals(deutw_fit$male$mod1))
deutw_peak_age <- as.integer(names(which.max(abs(deutw_fit$male$age_effects))))
deutw_peak_effect <- max(abs(deutw_fit$male$age_effects))
deutw_male_mean <- mean(abs(deutw_fit$male$age_effects))
deutw_female_mean <- mean(abs(deutw_fit$female$age_effects))

tibble(
  Measure = c(
    "Onset year (data-driven)",
    "AIC (baseline)", "AIC (full model)", "AIC improvement",
    "RMSE (baseline)", "RMSE (full model)", "RMSE reduction",
    "Peak age", "Peak impulse (log₁₀ units)",
    "Mean male impulse", "Mean female impulse", "M/F ratio"
  ),
  Value = c(
    as.character(deutw_best_onset),
    sprintf("%.0f", AIC(deutw_fit$male$mod0)),
    sprintf("%.0f", AIC(deutw_fit$male$mod1)),
    sprintf("%.0f", AIC(deutw_fit$male$mod0) - AIC(deutw_fit$male$mod1)),
    sprintf("%.4f", deutw_rms0),
    sprintf("%.4f", deutw_rms1),
    sprintf("%.1f%%", 100 * (1 - deutw_rms1 / deutw_rms0)),
    as.character(deutw_peak_age),
    sprintf("%.3f", deutw_peak_effect),
    sprintf("%.3f", deutw_male_mean),
    sprintf("%.3f", deutw_female_mean),
    sprintf("%.2f", deutw_male_mean / deutw_female_mean)
  )
) |>
  knitr::kable()
```

---

### Systematic comparison across all HMD countries {#sec-results-scan}

The impulse-decay model is fitted to every HMD country with sufficient data. @fig-scan-ranking shows the AIC improvement from adding the impulse-decay term, and @fig-decay-profile shows the AIC profile over decay rates for selected countries.

```{r}
#| label: scan-all-countries
#| include: false

# Fit impulse-decay model to each country and compute AIC improvement + sex ratio
scan_results <- map_dfr(unique(hmd$country_code), function(code) {
  country_data <- hmd |> filter(country_code == code)

  # Helper: fit impulse-decay for one sex
  fit_sex <- function(sex_val) {
    sex_data <- country_data |>
      filter(sex == sex_val, !is.na(deaths), !is.na(exposure),
             age >= AGE_MIN, age <= AGE_MAX) |>
      mutate(lmr = calc_lmr(deaths, exposure))

    if (nrow(sex_data) < 500) return(list(data = NULL, mod0 = NULL, mod1 = NULL))

    min_yr <- min(sex_data$year)
    model_data <- sex_data |>
      mutate(
        yr_pre_38 = year <= 1938,
        early_phase = ifelse(yr_pre_38, year - min_yr, 0),
        impr_phase = year >= 1939 & year <= 1955,
        yr_since_imp = ifelse(impr_phase, year - 1939, 0),
        post_55 = year > 1955,
        yr_since_imp2 = ifelse(post_55, year - 1955, 0),
        trbls_kpc = troubles_decay(year, OPTIMAL_DECAY_K)
      )

    mod0 <- tryCatch(
      lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
           yr_since_imp + yr_since_imp2), data = model_data),
      error = function(e) NULL
    )
    mod1 <- tryCatch(
      lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
           yr_since_imp + yr_since_imp2 + trbls_kpc), data = model_data),
      error = function(e) NULL
    )

    list(data = model_data, mod0 = mod0, mod1 = mod1)
  }

  male <- fit_sex("male")
  female <- fit_sex("female")

  if (is.null(male$mod0) || is.null(male$mod1)) return(NULL)

  # Extract mean absolute impulse-decay coefficient for each sex
  get_mean_impulse <- function(mod) {
    if (is.null(mod)) return(NA_real_)
    coefs <- coef(mod)
    trbls <- coefs[grep("trbls_kpc", names(coefs))]
    base <- trbls[1]
    inters <- if (length(trbls) > 1) trbls[-1] else numeric(0)
    age_effects <- c(base, base + inters)
    mean(abs(age_effects))
  }

  male_impulse <- get_mean_impulse(male$mod1)
  female_impulse <- get_mean_impulse(female$mod1)
  mf_ratio <- if (!is.na(female_impulse) && female_impulse > 0) {
    male_impulse / female_impulse
  } else {
    NA_real_
  }

  tibble(
    country_code = code,
    aic_baseline = AIC(male$mod0),
    aic_full = AIC(male$mod1),
    aic_improvement = AIC(male$mod0) - AIC(male$mod1),
    rms_baseline = rms(residuals(male$mod0)),
    rms_full = rms(residuals(male$mod1)),
    rms_reduction = 1 - rms(residuals(male$mod1)) / rms(residuals(male$mod0)),
    male_impulse = male_impulse,
    female_impulse = female_impulse,
    mf_ratio = mf_ratio
  )
})

country_labels <- c(
  GBR_NIR = "N. Ireland", GBR_NP = "UK", GBR_SCO = "Scotland",
  GBRCENW = "England & Wales (civ)", GBRTENW = "England & Wales",
  AUT = "Austria", BEL = "Belgium", CHE = "Switzerland",
  DEUTNP = "Germany", DEUTE = "East Germany", DEUTW = "West Germany",
  FRATNP = "France", FRACNP = "France (civ)", IRL = "Ireland",
  LUX = "Luxembourg", NLD = "Netherlands",
  ESP = "Spain", GRC = "Greece", ITA = "Italy", PRT = "Portugal",
  DNK = "Denmark", FIN = "Finland", ISL = "Iceland",
  NOR = "Norway", SWE = "Sweden",
  BGR = "Bulgaria", CZE = "Czechia", HRV = "Croatia",
  HUN = "Hungary", POL = "Poland", SVK = "Slovakia", SVN = "Slovenia",
  EST = "Estonia", LTU = "Lithuania", LVA = "Latvia",
  BLR = "Belarus", RUS = "Russia", UKR = "Ukraine",
  ISR = "Israel", CAN = "Canada", CHL = "Chile", USA = "USA",
  AUS = "Australia", JPN = "Japan", KOR = "South Korea",
  NZL_MA = "NZ Māori", NZL_NM = "NZ non-Māori", NZL_NP = "New Zealand",
  TWN = "Taiwan"
)

scan_results <- scan_results |>
  mutate(country = coalesce(country_labels[country_code], country_code)) |>
  arrange(desc(aic_improvement))
```

```{r}
#| label: fig-scan-ranking
#| fig-cap: "AIC improvement from adding the impulse-decay term (onset 1972, k = 0.097) across all comparator countries. Northern Ireland shows by far the largest improvement, as expected. Countries that experienced WWII on their territory (France, Germany, Belgium, Netherlands, Austria) show moderate improvements, reflecting war-related mortality disruptions."
#| fig-width: 8
#| fig-height: 6

ggplot(scan_results,
       aes(x = reorder(country, aic_improvement), y = aic_improvement)) +
  geom_col(aes(fill = aic_improvement > 50), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "grey60")) +
  coord_flip() +
  labs(x = NULL, y = "AIC improvement (baseline − full model)",
       subtitle = "Impulse-decay term: onset 1972, half-life 6.8 years") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-decay-profile
#| fig-cap: "AIC profile over decay rates for selected countries. The clear minimum for Northern Ireland at k ≈ 0.097 (half-life ~6.8 years) confirms a well-identified conflict signature. Countries without civil conflicts show flat or absent profiles."
#| fig-width: 8
#| fig-height: 6

# Profile over decay rates for selected countries
profile_countries <- c("GBR_NIR", "GBR_SCO", "GBRTENW", "IRL", "FRATNP")
decay_grid <- seq(0.01, 0.5, by = 0.01)

decay_profiles <- map_dfr(profile_countries, function(code) {
  country_data <- hmd |> filter(country_code == code)

  map_dfr(decay_grid, function(k) {
    model_data <- country_data |>
      filter(sex == "male", !is.na(deaths), !is.na(exposure)) |>
      prepare_model_data(sex_filter = "male",
                         age_range = c(AGE_MIN, AGE_MAX), k = k)

    mod <- tryCatch(fit_troubles_model(model_data), error = function(e) NULL)
    if (is.null(mod)) return(NULL)

    tibble(country_code = code, decay_rate = k, aic = AIC(mod))
  })
}) |>
  mutate(country = country_labels[country_code])

# Normalise AIC within each country for comparison
decay_profiles <- decay_profiles |>
  group_by(country) |>
  mutate(delta_aic = aic - min(aic)) |>
  ungroup()

ggplot(decay_profiles, aes(x = decay_rate, y = delta_aic, colour = country)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Decay rate (k)", y = "ΔAIC (relative to best fit)",
       colour = "Country") +
  scale_x_continuous(sec.axis = sec_axis(~ log(0.5) / log(1 - .),
                                          name = "Half-life (years)",
                                          breaks = c(1, 2, 5, 10, 20, 50))) +
  coord_cartesian(ylim = c(0, 200)) +
  theme_minimal(base_size = 12)
```

#### Sex-ratio diagnostic

The ratio of male to female impulse-decay coefficients characterises the nature of each detected disruption. @fig-sex-ratio plots this ratio against the AIC improvement (detection strength) for all countries in the scan.

```{r}
#| label: fig-sex-ratio
#| fig-cap: "Male-to-female impulse ratio versus detection strength across all HMD countries. Countries above the dashed line (ratio = 1) show predominantly male excess mortality. Northern Ireland's high ratio reflects combatant/sectarian violence concentrated among young men. Countries that experienced total war on their territory (e.g. France, Belgium) show lower ratios, reflecting civilian exposure in both sexes. Negative controls (Sweden, Switzerland) cluster near the origin."
#| fig-width: 9
#| fig-height: 7

scan_labelled <- scan_results |>
  filter(!is.na(mf_ratio), is.finite(mf_ratio))

ggplot(scan_labelled,
       aes(x = aic_improvement, y = mf_ratio, label = country)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_point(aes(size = male_impulse), alpha = 0.7, colour = "steelblue") +
  geom_text(aes(label = country), size = 2.5, vjust = -0.8,
            check_overlap = TRUE) +
  scale_size_continuous(name = "Male impulse\n(mean |coef|)", range = c(1, 6)) +
  labs(x = "AIC improvement (detection strength)",
       y = "Male / Female impulse ratio") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: tbl-sex-ratio
#| tbl-cap: "Male-to-female impulse ratio for countries with detectable disruptions (AIC improvement > 50). A high ratio indicates predominantly male excess mortality, consistent with combatant or sectarian violence; a ratio near 1 indicates both-sex exposure, consistent with total war, occupation, or non-conflict disruption."

scan_results |>
  filter(aic_improvement > 50, !is.na(mf_ratio)) |>
  arrange(desc(mf_ratio)) |>
  transmute(
    Country = country,
    `AIC improvement` = sprintf("%.0f", aic_improvement),
    `Male impulse` = sprintf("%.3f", male_impulse),
    `Female impulse` = sprintf("%.3f", female_impulse),
    `M/F ratio` = sprintf("%.2f", mf_ratio)
  ) |>
  knitr::kable()
```

---

### Summary of estimates {#sec-results-summary}

@tbl-headline presents the paper's principal output for the Northern Ireland validation case.

```{r}
#| label: tbl-headline
#| tbl-cap: "Conflict dynamics and excess mortality estimates for Northern Ireland from all-cause data. All estimates refer to males aged 15--45. Northern Ireland validated against independent tallies."

# Compute key summary statistics
five_yr_excess <- excess |>
  filter(year >= 1972, year <= 1976) |>
  summarise(total = sum(excess, na.rm = TRUE)) |>
  pull(total) |>
  round()

cumulative_to_date <- round(max(excess_by_year$cumulative))

# Asymptotic total (geometric series: sum = first_year_excess / k)
first_year_excess <- excess |>
  filter(year == 1972) |>
  summarise(total = sum(excess, na.rm = TRUE)) |>
  pull(total)
asymptotic_total <- round(first_year_excess / k_opt)

tibble(
  Measure = c(
    "Onset year",
    "Decay rate (k)",
    "Half-life (years)",
    "Peak age",
    "Peak impulse (log₁₀ units)",
    "Time to 5% intensity",
    "Excess deaths (first 5 years)",
    paste0("Cumulative excess to ", max(excess_by_year$year)),
    "Expected total (asymptotic)",
    "Independent tally (Smyth 1998)",
    "Ratio (model / tally)"
  ),
  Value = c(
    as.character(TROUBLES_START_YEAR),
    sprintf("%.4f", k_opt),
    sprintf("%.1f", halflife),
    as.character(peak_age),
    sprintf("%.3f", peak_effect),
    sprintf("%.0f years (~%d)", time_to_5pct, round(TROUBLES_START_YEAR + time_to_5pct)),
    format(five_yr_excess, big.mark = ","),
    format(cumulative_to_date, big.mark = ","),
    format(asymptotic_total, big.mark = ","),
    "3,598",
    sprintf("%.2f", cumulative_to_date / 3598)
  )
) |>
  knitr::kable()
```

The table characterises the conflict along two axes: **intensity** (impulse size, 5-year excess) and **persistence** (half-life, time to 5%). Northern Ireland exemplifies a lower-intensity, long-persistence conflict: a conflict of moderate initial intensity that was sustained endogenously over decades through tit-for-tat dynamics.

The "expected total" column exploits the parametric form of the model. Because the excess decays geometrically, the asymptotic total is calculable as an infinite geometric series. The difference between "cumulative to date" and "expected total" indicates how much residual excess mortality is still accruing — a quantity that is invisible without a model but is relevant to ongoing policy and post-conflict health planning. For Northern Ireland, the model implies the Troubles were still contributing measurable (if small) excess mortality well into the 2000s, long after the Good Friday Agreement.
