### Northern Ireland: validation case {#sec-results-ni}

Northern Ireland serves as the validation case: the Troubles are well-documented, independent conflict-death tallies exist, and HMD data quality is high. The model's estimates can be directly compared against these tallies to assess calibration.

```{r}
#| label: load-data
#| include: false

# Load HMD data
hmd <- read_csv("data/processed/hmd_counts.csv", show_col_types = FALSE)
ni <- hmd |> filter(country_code == "GBR_NIR")

# Prepare NI male data for modelling
ni_male <- ni |>
  filter(sex == "male") |>
  filter(!is.na(deaths), !is.na(exposure)) |>
  prepare_model_data(sex_filter = "male", age_range = c(AGE_MIN, AGE_MAX), k = OPTIMAL_DECAY_K)

# Fit the full model (model 5 from original analysis)
mod_full <- fit_troubles_model(ni_male)

# Predictions with Troubles
ni_male_preds <- ni_male |>
  modelr::add_predictions(mod_full, var = "pred_full")

# Counterfactual predictions (no Troubles)
ni_male_counter <- ni_male |>
  mutate(trbls_kpc = 0) |>
  broom::augment(mod_full, newdata = _) |>
  select(year, age, pred_counter = .fitted)

# Combine
ni_decomp <- ni_male_preds |>
  left_join(ni_male_counter, by = c("year", "age")) |>
  mutate(impulse_decay = pred_full - pred_counter)
```

#### Visual decomposition

@fig-ni-decomposition presents the four-panel decomposition for males aged `r AGE_MIN`--`r AGE_MAX` in Northern Ireland, 1922--2023.

```{r}
#| label: fig-ni-decomposition
#| fig-cap: "Decomposition of the mortality surface for males aged 15--45 in Northern Ireland. (A) Observed log₁₀ mortality rates. (B) Full model: smooth baseline plus impulse-decay component. (C) Counterfactual: smooth baseline with impulse-decay turned off. (D) Impulse-decay component alone (B minus C), representing estimated conflict-attributable excess mortality."
#| fig-width: 10
#| fig-height: 10

lmr_range <- range(c(ni_decomp$lmr, ni_decomp$pred_full, ni_decomp$pred_counter), na.rm = TRUE)

panel_a <- ggplot(ni_decomp, aes(x = year, y = age, fill = lmr)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "A: Observed", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_b <- ggplot(ni_decomp, aes(x = year, y = age, fill = pred_full)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "B: Full model", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_c <- ggplot(ni_decomp, aes(x = year, y = age, fill = pred_counter)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
                       limits = lmr_range, name = "log₁₀(m)") +
  labs(title = "C: Counterfactual", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

panel_d <- ggplot(ni_decomp |> filter(year >= 1960),
                  aes(x = year, y = age, fill = impulse_decay)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "firebrick",
                       midpoint = 0, name = "Δ log₁₀(m)") +
  labs(title = "D: Impulse-decay component", x = "Year", y = "Age") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)

patchwork <- (panel_a + panel_b) / (panel_c + panel_d)
patchwork
```

Panel A shows the raw data, including the visible band of elevated mortality in young adult males after 1972. Panel B shows the model's stylised representation — the reader can assess fit by comparing A and B. Panel C shows the counterfactual: what the surface would have looked like had the Troubles not occurred. Panel D isolates the impulse-decay component itself — the conflict's estimated demographic footprint.

#### Parameter estimates

```{r}
#| label: tbl-ni-params
#| tbl-cap: "Estimated impulse-decay parameters for Northern Ireland"

# Optimal decay rate (pre-computed via AIC optimisation)
optimal <- find_optimal_decay(ni, criterion = "AIC")
k_opt <- optimal$minimum
halflife <- log(0.5) / log(1 - k_opt)
time_to_5pct <- log(0.05) / log(1 - k_opt)

# Extract Troubles coefficients by age
coefs <- coef(mod_full)
trbls_coefs <- coefs[grep("trbls_kpc", names(coefs))]
# First is the base (age 15), rest are interactions
base_effect <- trbls_coefs[1]
interaction_effects <- trbls_coefs[-1]
age_effects <- c(base_effect, base_effect + interaction_effects)
names(age_effects) <- AGE_MIN:AGE_MAX
peak_age <- as.integer(names(which.max(abs(age_effects))))
peak_effect <- max(abs(age_effects))

tibble(
  Parameter = c("Onset year", "Decay rate (k)", "Half-life (years)",
                "Peak age", "Peak impulse (log₁₀ units)",
                "Time to 5% intensity (years)"),
  Estimate = c(
    as.character(TROUBLES_START_YEAR),
    sprintf("%.4f", k_opt),
    sprintf("%.1f", halflife),
    as.character(peak_age),
    sprintf("%.3f", peak_effect),
    sprintf("%.0f (~%d)", time_to_5pct, round(TROUBLES_START_YEAR + time_to_5pct))
  )
) |>
  knitr::kable()
```

#### Cumulative excess deaths

@fig-ni-cumulative shows the cumulative estimated excess deaths over time, derived by summing the impulse-decay component (panel D) across ages, weighted by population exposure.

```{r}
#| label: fig-ni-cumulative
#| fig-cap: "Cumulative estimated excess male deaths (ages 15--45) attributable to the Troubles. Dashed line: approximate cumulative directly-attributed conflict deaths from Smyth (1998)."

excess <- estimate_excess_deaths(ni, mod_full, ni_male)

excess_by_year <- excess |>
  filter(year >= 1970) |>
  group_by(year) |>
  summarise(annual_excess = sum(excess, na.rm = TRUE), .groups = "drop") |>
  mutate(cumulative = cumsum(annual_excess))

total_excess <- round(max(excess_by_year$cumulative))

# Smyth (1998) tally: ~3,598 total; roughly scaled to males 15-45
smyth_total <- 3598

ggplot(excess_by_year, aes(x = year, y = cumulative)) +
  geom_line(linewidth = 1, colour = "firebrick") +
  geom_hline(yintercept = smyth_total, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 2010, y = smyth_total + 100,
           label = paste0("Smyth (1998): ", format(smyth_total, big.mark = ",")),
           size = 3.5, colour = "grey40") +
  annotate("text", x = 2010, y = total_excess - 200,
           label = paste0("Model estimate: ", format(total_excess, big.mark = ",")),
           size = 3.5, colour = "firebrick") +
  labs(x = "Year", y = "Cumulative excess deaths",
       subtitle = paste0("Males aged ", AGE_MIN, "-", AGE_MAX,
                         "; model captures ~",
                         round(100 * total_excess / smyth_total), "% of tallied deaths")) +
  theme_minimal(base_size = 12)
```

The comparison with the independent tally from @smyth1998 provides a direct calibration check. Our model, using only all-cause mortality data, is expected to capture approximately 75% of the directly-attributed deaths — the remainder reflecting mortality displacement effects, emigration as a competing risk, and the restriction to males aged `r AGE_MIN`--`r AGE_MAX` (see @sec-comparison).

#### Troubles effect by age and sex

As a specificity check, the same model is applied to the female mortality surface. @fig-ni-sex-comparison shows the impulse-decay coefficient by age for both sexes.

```{r}
#| label: fig-ni-sex-comparison
#| fig-cap: "Impulse-decay coefficient by age for males and females in Northern Ireland. The conflict signature is concentrated in young adult males, with a peak near age 18--20. The female surface shows negligible or no conflict effect, consistent with the sex-specific nature of sectarian violence."

# Male coefficients
male_effects <- tibble(
  age = AGE_MIN:AGE_MAX,
  effect = as.numeric(age_effects),
  sex = "Male"
)

# Female model
ni_female <- ni |>
  filter(sex == "female") |>
  filter(!is.na(deaths), !is.na(exposure)) |>
  prepare_model_data(sex_filter = "female", age_range = c(AGE_MIN, AGE_MAX), k = OPTIMAL_DECAY_K)

mod_female <- fit_troubles_model(ni_female)
coefs_f <- coef(mod_female)
trbls_coefs_f <- coefs_f[grep("trbls_kpc", names(coefs_f))]
base_f <- trbls_coefs_f[1]
inter_f <- trbls_coefs_f[-1]
age_effects_f <- c(base_f, base_f + inter_f)

female_effects <- tibble(
  age = AGE_MIN:AGE_MAX,
  effect = as.numeric(age_effects_f),
  sex = "Female"
)

bind_rows(male_effects, female_effects) |>
  ggplot(aes(x = age, y = effect, colour = sex, shape = sex)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 18, linetype = "dotted", alpha = 0.5) +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Age", y = "Impulse-decay coefficient",
       colour = "Sex", shape = "Sex") +
  theme_minimal(base_size = 12)
```

#### Residual surface

```{r}
#| label: fig-ni-residuals
#| fig-cap: "Residual surface (predicted minus observed) for the full model fitted to Northern Ireland males aged 15--45. The absence of structured residual patterns after 1972 indicates that the impulse-decay component adequately captures the conflict signature."

ni_decomp |>
  mutate(residual = pred_full - lmr) |>
  ggplot(aes(x = year, y = age, fill = residual)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0,
                       limits = c(-0.6, 0.6),
                       oob = scales::squish,
                       name = "Residual") +
  labs(x = "Year", y = "Age") +
  theme_minimal(base_size = 12) +
  coord_fixed(ratio = 1)
```

#### Model comparison

```{r}
#| label: tbl-ni-model-comparison
#| tbl-cap: "Model comparison for Northern Ireland males. The inclusion of the impulse-decay term (Model 5) substantially improves fit, with a ~19% reduction in root mean square error."

# Fit models without Troubles for comparison
mod_no_troubles <- lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
                        yr_since_imp + yr_since_imp2),
                      data = ni_male)

rms_no_troubles <- rms(residuals(mod_no_troubles))
rms_full <- rms(residuals(mod_full))

tibble(
  Model = c("Baseline (no Troubles term)", "Full (with impulse-decay)"),
  AIC = c(AIC(mod_no_troubles), AIC(mod_full)),
  BIC = c(BIC(mod_no_troubles), BIC(mod_full)),
  RMSE = c(rms_no_troubles, rms_full),
  `RMSE reduction` = c("—", sprintf("%.1f%%", 100 * (1 - rms_full / rms_no_troubles)))
) |>
  mutate(across(c(AIC, BIC), \(x) sprintf("%.0f", x)),
         RMSE = sprintf("%.4f", RMSE)) |>
  knitr::kable()
```

---

### Systematic comparison across UK and Western Europe {#sec-results-scan}

The impulse-decay model is fitted to each comparator country. @fig-scan-ranking shows the AIC improvement from adding the impulse-decay term, and @fig-decay-profile shows the AIC profile over decay rates for each country.

```{r}
#| label: scan-all-countries
#| include: false

# Fit impulse-decay model to each country and compute AIC improvement
scan_results <- map_dfr(unique(hmd$country_code), function(code) {
  country_data <- hmd |> filter(country_code == code)

  # Check we have enough data
  male_data <- country_data |>
    filter(sex == "male", !is.na(deaths), !is.na(exposure),
           age >= AGE_MIN, age <= AGE_MAX) |>
    mutate(lmr = calc_lmr(deaths, exposure))

  if (nrow(male_data) < 500) return(NULL)

  # Prepare data with NI-style phases (adapt to each country's year range)
  min_yr <- min(male_data$year)
  model_data <- male_data |>
    mutate(
      yr_pre_38 = year <= 1938,
      early_phase = ifelse(yr_pre_38, year - min_yr, 0),
      impr_phase = year >= 1939 & year <= 1955,
      yr_since_imp = ifelse(impr_phase, year - 1939, 0),
      post_55 = year > 1955,
      yr_since_imp2 = ifelse(post_55, year - 1955, 0),
      trbls_kpc = troubles_decay(year, OPTIMAL_DECAY_K)
    )

  # Baseline model (no Troubles)
  mod0 <- tryCatch(
    lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
         yr_since_imp + yr_since_imp2), data = model_data),
    error = function(e) NULL
  )

  # Full model (with Troubles decay)
  mod1 <- tryCatch(
    lm(lmr ~ as.factor(age) * (yr_pre_38 + early_phase + impr_phase +
         yr_since_imp + yr_since_imp2 + trbls_kpc), data = model_data),
    error = function(e) NULL
  )

  if (is.null(mod0) || is.null(mod1)) return(NULL)

  tibble(
    country_code = code,
    aic_baseline = AIC(mod0),
    aic_full = AIC(mod1),
    aic_improvement = AIC(mod0) - AIC(mod1),
    rms_baseline = rms(residuals(mod0)),
    rms_full = rms(residuals(mod1)),
    rms_reduction = 1 - rms(residuals(mod1)) / rms(residuals(mod0))
  )
})

country_labels <- c(
  GBR_NIR = "N. Ireland", GBR_SCO = "Scotland", GBRTENW = "England & Wales",
  IRL = "Ireland", FRATNP = "France", BEL = "Belgium",
  LUX = "Luxembourg", NLD = "Netherlands", DEUTNP = "Germany",
  CHE = "Switzerland", AUT = "Austria"
)

scan_results <- scan_results |>
  mutate(country = country_labels[country_code]) |>
  arrange(desc(aic_improvement))
```

```{r}
#| label: fig-scan-ranking
#| fig-cap: "AIC improvement from adding the impulse-decay term (onset 1972, k = 0.097) across all comparator countries. Northern Ireland shows by far the largest improvement, as expected. Countries that experienced WWII on their territory (France, Germany, Belgium, Netherlands, Austria) show moderate improvements, reflecting war-related mortality disruptions."
#| fig-width: 8
#| fig-height: 6

ggplot(scan_results,
       aes(x = reorder(country, aic_improvement), y = aic_improvement)) +
  geom_col(aes(fill = aic_improvement > 50), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "grey60")) +
  coord_flip() +
  labs(x = NULL, y = "AIC improvement (baseline − full model)",
       subtitle = "Impulse-decay term: onset 1972, half-life 6.8 years") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-decay-profile
#| fig-cap: "AIC profile over decay rates for selected countries. The clear minimum for Northern Ireland at k ≈ 0.097 (half-life ~6.8 years) confirms a well-identified conflict signature. Countries without civil conflicts show flat or absent profiles."
#| fig-width: 8
#| fig-height: 6

# Profile over decay rates for selected countries
profile_countries <- c("GBR_NIR", "GBR_SCO", "GBRTENW", "IRL", "FRATNP")
decay_grid <- seq(0.01, 0.5, by = 0.01)

decay_profiles <- map_dfr(profile_countries, function(code) {
  country_data <- hmd |> filter(country_code == code)

  map_dfr(decay_grid, function(k) {
    model_data <- country_data |>
      filter(sex == "male", !is.na(deaths), !is.na(exposure)) |>
      prepare_model_data(sex_filter = "male",
                         age_range = c(AGE_MIN, AGE_MAX), k = k)

    mod <- tryCatch(fit_troubles_model(model_data), error = function(e) NULL)
    if (is.null(mod)) return(NULL)

    tibble(country_code = code, decay_rate = k, aic = AIC(mod))
  })
}) |>
  mutate(country = country_labels[country_code])

# Normalise AIC within each country for comparison
decay_profiles <- decay_profiles |>
  group_by(country) |>
  mutate(delta_aic = aic - min(aic)) |>
  ungroup()

ggplot(decay_profiles, aes(x = decay_rate, y = delta_aic, colour = country)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Decay rate (k)", y = "ΔAIC (relative to best fit)",
       colour = "Country") +
  scale_x_continuous(sec.axis = sec_axis(~ log(0.5) / log(1 - .),
                                          name = "Half-life (years)",
                                          breaks = c(1, 2, 5, 10, 20, 50))) +
  coord_cartesian(ylim = c(0, 200)) +
  theme_minimal(base_size = 12)
```

---

### Summary of estimates {#sec-results-summary}

@tbl-headline presents the paper's principal output for the Northern Ireland validation case.

```{r}
#| label: tbl-headline
#| tbl-cap: "Conflict dynamics and excess mortality estimates for Northern Ireland from all-cause data. All estimates refer to males aged 15--45. Northern Ireland validated against independent tallies."

# Compute key summary statistics
five_yr_excess <- excess |>
  filter(year >= 1972, year <= 1976) |>
  summarise(total = sum(excess, na.rm = TRUE)) |>
  pull(total) |>
  round()

cumulative_to_date <- round(max(excess_by_year$cumulative))

# Asymptotic total (geometric series: sum = first_year_excess / k)
first_year_excess <- excess |>
  filter(year == 1972) |>
  summarise(total = sum(excess, na.rm = TRUE)) |>
  pull(total)
asymptotic_total <- round(first_year_excess / k_opt)

tibble(
  Measure = c(
    "Onset year",
    "Decay rate (k)",
    "Half-life (years)",
    "Peak age",
    "Peak impulse (log₁₀ units)",
    "Time to 5% intensity",
    "Excess deaths (first 5 years)",
    paste0("Cumulative excess to ", max(excess_by_year$year)),
    "Expected total (asymptotic)",
    "Independent tally (Smyth 1998)",
    "Ratio (model / tally)"
  ),
  Value = c(
    as.character(TROUBLES_START_YEAR),
    sprintf("%.4f", k_opt),
    sprintf("%.1f", halflife),
    as.character(peak_age),
    sprintf("%.3f", peak_effect),
    sprintf("%.0f years (~%d)", time_to_5pct, round(TROUBLES_START_YEAR + time_to_5pct)),
    format(five_yr_excess, big.mark = ","),
    format(cumulative_to_date, big.mark = ","),
    format(asymptotic_total, big.mark = ","),
    "3,598",
    sprintf("%.2f", cumulative_to_date / 3598)
  )
) |>
  knitr::kable()
```

The table characterises the conflict along two axes: **intensity** (impulse size, 5-year excess) and **persistence** (half-life, time to 5%). Northern Ireland exemplifies a lower-intensity, long-persistence conflict: a conflict of moderate initial intensity that was sustained endogenously over decades through tit-for-tat dynamics.

The "expected total" column exploits the parametric form of the model. Because the excess decays geometrically, the asymptotic total is calculable as an infinite geometric series. The difference between "cumulative to date" and "expected total" indicates how much residual excess mortality is still accruing — a quantity that is invisible without a model but is relevant to ongoing policy and post-conflict health planning. For Northern Ireland, the model implies the Troubles were still contributing measurable (if small) excess mortality well into the 2000s, long after the Good Friday Agreement.
