## Appendix A: Detailed results for all HMD populations {#sec-appendix-all .appendix}

This appendix presents the impulse-decay model results for all `r length(ALL_HMD_CODES)` populations in the Human Mortality Database. For each population, four panels are shown:

1. **Raw death counts** by sex (ages `r AGE_MIN`--`r AGE_MAX`) with a vertical dashed line at the NI-calibrated onset year (1972)
2. **Impulse-decay coefficient by age** for males and females
3. **Residual surface** for the male model
4. **Summary statistics** including AIC improvement, RMSE reduction, peak age, and male/female impulse ratio

Populations are grouped by region and ordered by detection strength (AIC improvement). The simplified baseline (quadratic year trend) is used for all populations to ensure comparability.

```{r}
#| label: appendix-setup
#| include: false

# Country metadata for grouping and labelling
country_meta <- tribble(
  ~code, ~name, ~region,
  "GBR_NIR", "Northern Ireland", "UK",
  "GBR_NP", "United Kingdom", "UK",
  "GBR_SCO", "Scotland", "UK",
  "GBRCENW", "England & Wales (civilian)", "UK",
  "GBRTENW", "England & Wales (total)", "UK",
  "AUT", "Austria", "Western Europe",
  "BEL", "Belgium", "Western Europe",
  "CHE", "Switzerland", "Western Europe",
  "DEUTNP", "Germany (unified)", "Western Europe",
  "DEUTE", "East Germany", "Western Europe",
  "DEUTW", "West Germany", "Western Europe",
  "FRATNP", "France (total)", "Western Europe",
  "FRACNP", "France (civilian)", "Western Europe",
  "IRL", "Ireland", "Western Europe",
  "LUX", "Luxembourg", "Western Europe",
  "NLD", "Netherlands", "Western Europe",
  "ESP", "Spain", "Southern Europe",
  "GRC", "Greece", "Southern Europe",
  "ITA", "Italy", "Southern Europe",
  "PRT", "Portugal", "Southern Europe",
  "DNK", "Denmark", "Scandinavia",
  "FIN", "Finland", "Scandinavia",
  "ISL", "Iceland", "Scandinavia",
  "NOR", "Norway", "Scandinavia",
  "SWE", "Sweden", "Scandinavia",
  "BGR", "Bulgaria", "Central & Eastern Europe",
  "CZE", "Czechia", "Central & Eastern Europe",
  "HRV", "Croatia", "Central & Eastern Europe",
  "HUN", "Hungary", "Central & Eastern Europe",
  "POL", "Poland", "Central & Eastern Europe",
  "SVK", "Slovakia", "Central & Eastern Europe",
  "SVN", "Slovenia", "Central & Eastern Europe",
  "EST", "Estonia", "Baltic States",
  "LTU", "Lithuania", "Baltic States",
  "LVA", "Latvia", "Baltic States",
  "BLR", "Belarus", "Former Soviet Union",
  "RUS", "Russia", "Former Soviet Union",
  "UKR", "Ukraine", "Former Soviet Union",
  "ISR", "Israel", "Middle East",
  "CAN", "Canada", "Americas",
  "CHL", "Chile", "Americas",
  "USA", "United States", "Americas",
  "AUS", "Australia", "Asia-Pacific",
  "JPN", "Japan", "Asia-Pacific",
  "KOR", "South Korea", "Asia-Pacific",
  "NZL_MA", "New Zealand (Māori)", "Asia-Pacific",
  "NZL_NM", "New Zealand (non-Māori)", "Asia-Pacific",
  "NZL_NP", "New Zealand (total)", "Asia-Pacific",
  "TWN", "Taiwan", "Asia-Pacific"
)

# Fit all populations with onset year scanning
fit_all_populations <- function(hmd_data, country_meta,
                                 k = OPTIMAL_DECAY_K,
                                 age_range = c(AGE_MIN, AGE_MAX)) {

  results <- map(country_meta$code, function(code) {
    country_data <- hmd_data |> filter(country_code == code, !is.na(year))

    # First: scan for optimal onset year using males
    male_scan <- country_data |>
      filter(sex == "male", !is.na(deaths), !is.na(exposure),
             age >= age_range[1], age <= age_range[2]) |>
      mutate(lmr = calc_lmr(deaths, exposure),
             yr_centred = year - median(year))

    if (nrow(male_scan) < 200) return(NULL)

    yr_range <- range(male_scan$year)
    onset_candidates <- seq(yr_range[1] + 5, yr_range[2] - 5)

    onset_aics <- map_dfr(onset_candidates, function(t0) {
      d <- male_scan |> mutate(impulse = ifelse(year < t0, 0, (1-k)^(year-t0)))
      mod <- tryCatch(
        lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2) + impulse), data = d),
        error = function(e) NULL)
      if (is.null(mod)) return(NULL)
      tibble(onset = t0, aic = AIC(mod))
    })

    best_onset <- onset_aics |> slice_min(aic) |> pull(onset) |> first()
    onset <- best_onset

    fit_one_sex <- function(sex_val) {
      sex_data <- country_data |>
        filter(sex == sex_val, !is.na(deaths), !is.na(exposure),
               age >= age_range[1], age <= age_range[2]) |>
        mutate(
          lmr = calc_lmr(deaths, exposure),
          yr_centred = year - median(year),
          impulse = ifelse(year < onset, 0, (1 - k) ^ (year - onset))
        )

      if (nrow(sex_data) < 200) return(NULL)

      mod0 <- tryCatch(
        lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2)),
           data = sex_data),
        error = function(e) NULL
      )
      mod1 <- tryCatch(
        lm(lmr ~ as.factor(age) * (yr_centred + I(yr_centred^2) + impulse),
           data = sex_data),
        error = function(e) NULL
      )

      if (is.null(mod0) || is.null(mod1)) return(NULL)

      preds <- sex_data |>
        modelr::add_predictions(mod1, var = "pred_full")
      counter <- sex_data |>
        mutate(impulse = 0) |>
        broom::augment(mod1, newdata = _) |>
        select(year, age, pred_counter = .fitted)

      decomp <- preds |>
        left_join(counter, by = c("year", "age")) |>
        mutate(impulse_decay = pred_full - pred_counter,
               residual = pred_full - lmr)

      coefs <- coef(mod1)
      imp_coefs <- coefs[grep("impulse", names(coefs))]
      base_imp <- imp_coefs[1]
      inter_imp <- if (length(imp_coefs) > 1) imp_coefs[-1] else numeric(0)
      age_effects <- c(base_imp, base_imp + inter_imp)
      names(age_effects) <- age_range[1]:age_range[2]

      list(mod0 = mod0, mod1 = mod1, decomp = decomp,
           age_effects = age_effects, data = sex_data)
    }

    male <- fit_one_sex("male")
    female <- fit_one_sex("female")

    if (is.null(male)) return(NULL)

    male_impulse <- mean(abs(male$age_effects))
    female_impulse <- if (!is.null(female)) mean(abs(female$age_effects)) else NA_real_
    mf_ratio <- if (!is.na(female_impulse) && female_impulse > 0) {
      male_impulse / female_impulse
    } else NA_real_

    meta <- country_meta |> filter(code == !!code)

    list(
      code = code,
      name = meta$name,
      region = meta$region,
      male = male,
      female = female,
      best_onset = best_onset,
      onset_aics = onset_aics,
      aic_improvement = AIC(male$mod0) - AIC(male$mod1),
      rms_baseline = rms(residuals(male$mod0)),
      rms_full = rms(residuals(male$mod1)),
      rms_reduction = 1 - rms(residuals(male$mod1)) / rms(residuals(male$mod0)),
      male_impulse = male_impulse,
      female_impulse = female_impulse,
      mf_ratio = mf_ratio,
      peak_age = as.integer(names(which.max(abs(male$age_effects)))),
      peak_effect = max(abs(male$age_effects)),
      year_range = paste0(min(male$data$year), "--", max(male$data$year)),
      n_years = n_distinct(male$data$year)
    )
  })

  # Remove NULLs and sort by AIC improvement
  results <- compact(results)
  results[order(-sapply(results, \(x) x$aic_improvement))]
}

all_fits <- fit_all_populations(hmd, country_meta)
```

### Summary table

@tbl-appendix-summary presents summary statistics for all populations, ranked by detection strength.

```{r}
#| label: tbl-appendix-summary
#| tbl-cap: "Impulse-decay detection results for all HMD populations. Ranked by AIC improvement (detection strength). M/F ratio > 1 indicates predominantly male excess mortality."

summary_tbl <- map_dfr(all_fits, function(x) {
  tibble(
    Country = x$name,
    Region = x$region,
    Years = x$year_range,
    `Onset` = as.character(x$best_onset),
    `AIC impr.` = sprintf("%.0f", x$aic_improvement),
    `RMSE red.` = sprintf("%.1f%%", 100 * x$rms_reduction),
    `Peak age` = as.character(x$peak_age),
    `M impulse` = sprintf("%.3f", x$male_impulse),
    `F impulse` = sprintf("%.3f", x$female_impulse),
    `M/F ratio` = if (!is.na(x$mf_ratio)) sprintf("%.2f", x$mf_ratio) else "—"
  )
})

knitr::kable(summary_tbl)
```

### Population-level detail

```{r}
#| label: appendix-all-populations
#| results: asis
#| fig-width: 10
#| fig-height: 12
#| warning: false

for (fit in all_fits) {
  cat("\n\n#### ", fit$name, " (", fit$code, ")\n\n", sep = "")
  cat("**Region:** ", fit$region, " | ",
      "**Years:** ", fit$year_range, " | ",
      "**Optimal onset:** ", fit$best_onset, " | ",
      "**AIC improvement:** ", sprintf("%.0f", fit$aic_improvement), " | ",
      "**RMSE reduction:** ", sprintf("%.1f%%", 100 * fit$rms_reduction), " | ",
      "**M/F ratio:** ",
      if (!is.na(fit$mf_ratio)) sprintf("%.2f", fit$mf_ratio) else "—",
      "\n\n", sep = "")

  country_data <- hmd |> filter(country_code == fit$code, !is.na(year))

  # Panel 1: Raw counts
  p_raw <- country_data |>
    filter(!is.na(deaths), age >= AGE_MIN, age <= AGE_MAX) |>
    group_by(year, sex) |>
    summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
    mutate(sex = str_to_title(sex)) |>
    ggplot(aes(x = year, y = deaths, colour = sex)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_line(linewidth = 0.3, alpha = 0.4) +
    geom_vline(xintercept = 1972, linetype = "dashed", colour = "grey40") +
    scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
    labs(title = paste0("A: Raw deaths (ages ", AGE_MIN, "-", AGE_MAX, ")"),
         x = "Year", y = "Deaths", colour = "Sex") +
    theme_minimal(base_size = 10)

  # Panel 2: Sex comparison of impulse coefficients
  sex_effects <- tibble(
    age = AGE_MIN:AGE_MAX,
    effect = as.numeric(fit$male$age_effects),
    sex = "Male"
  )
  if (!is.null(fit$female)) {
    sex_effects <- bind_rows(sex_effects,
      tibble(age = AGE_MIN:AGE_MAX,
             effect = as.numeric(fit$female$age_effects),
             sex = "Female"))
  }

  p_sex <- ggplot(sex_effects, aes(x = age, y = effect, colour = sex)) +
    geom_point(size = 1.5) + geom_line(linewidth = 0.4) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
    labs(title = "B: Impulse coefficient by age",
         x = "Age", y = "Coefficient", colour = "Sex") +
    theme_minimal(base_size = 10)

  # Panel 3: Male Lexis residual surface
  p_resid <- fit$male$decomp |>
    ggplot(aes(x = year, y = age, fill = residual)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0, limits = c(-0.6, 0.6),
                         oob = scales::squish, name = "Resid.") +
    labs(title = "C: Residuals (male model)", x = "Year", y = "Age") +
    theme_minimal(base_size = 10) + coord_fixed(ratio = 1)

  # Panel 4: Impulse-decay component
  p_impulse <- fit$male$decomp |>
    ggplot(aes(x = year, y = age, fill = impulse_decay)) +
    geom_tile() +
    scale_fill_gradient2(low = "navyblue", mid = "white", high = "firebrick",
                         midpoint = 0, name = "Δ log₁₀(m)") +
    labs(title = "D: Impulse-decay component (male)", x = "Year", y = "Age") +
    theme_minimal(base_size = 10) + coord_fixed(ratio = 1)

  combined <- (p_raw + p_sex) / (p_resid + p_impulse) +
    patchwork::plot_annotation(
      title = paste0(fit$name, " (", fit$code, ")"),
      subtitle = paste0("AIC improvement: ", sprintf("%.0f", fit$aic_improvement),
                        " | M/F ratio: ",
                        if (!is.na(fit$mf_ratio)) sprintf("%.2f", fit$mf_ratio) else "—")
    )

  print(combined)
  cat("\n\n")
}
```
