## Appendix B: Bayesian estimation via rstanarm {#sec-appendix-bayesian .appendix}

This appendix presents Bayesian estimates of the impulse-decay model for all HMD populations, providing posterior credible intervals for the conflict effect parameters. Models are fitted using `rstanarm::stan_glm` [@rstanarm2024], which uses pre-compiled Stan code with Hamiltonian Monte Carlo (NUTS) sampling.

The Bayesian model uses a simplified parameterisation with continuous age effects (quadratic) rather than age-as-factor, enabling efficient estimation:

$$\text{lmr}_{a,t} = \beta_0 + \beta_1 a' + \beta_2 a'^2 + \beta_3 t' + \beta_4 t'^2 + (\gamma_0 + \gamma_1 a' + \gamma_2 a'^2) \cdot I(t) + \varepsilon_{a,t}$$

where $a' = a - \bar{a}$ (centred age), $t' = t - 1970$ (centred year), and $I(t)$ is the impulse-decay term at the data-driven optimal onset year. Default weakly informative priors are used throughout.

The key quantity of interest is $P(\gamma_0 > 0)$: the posterior probability that the impulse-decay term captures positive excess mortality. Values near 1 indicate strong evidence of a mortality disruption at the identified onset; values near 0.5 indicate no evidence.

```{r}
#| label: bayesian-setup
#| include: false

library(rstanarm)
options(mc.cores = parallel::detectCores())

# Fit Bayesian model for each country
bayes_results <- map_dfr(all_fits, function(fit) {
  country_data <- hmd |> filter(country_code == fit$code, !is.na(year))
  onset <- fit$best_onset

  fit_bayes_sex <- function(sex_val) {
    sex_data <- country_data |>
      filter(sex == sex_val, !is.na(deaths), !is.na(exposure),
             age >= AGE_MIN, age <= AGE_MAX) |>
      mutate(
        lmr = calc_lmr(deaths, exposure),
        age_c = age - mean(age),
        age_c2 = (age - mean(age))^2,
        yr_c = year - 1970,
        impulse = ifelse(year < onset, 0,
                         (1 - OPTIMAL_DECAY_K) ^ (year - onset))
      )

    if (nrow(sex_data) < 200) return(NULL)

    mod <- tryCatch(
      stan_glm(
        lmr ~ age_c + age_c2 + yr_c + I(yr_c^2) + impulse +
               age_c:impulse + age_c2:impulse,
        data = sex_data,
        chains = 4, iter = 2000, seed = 42, refresh = 0
      ),
      error = function(e) NULL
    )

    if (is.null(mod)) return(NULL)

    draws <- as.data.frame(mod)
    impulse_draws <- draws[["impulse"]]

    tibble(
      median = median(impulse_draws),
      lower = quantile(impulse_draws, 0.025),
      upper = quantile(impulse_draws, 0.975),
      p_positive = mean(impulse_draws > 0)
    )
  }

  male_res <- fit_bayes_sex("male")
  female_res <- fit_bayes_sex("female")

  if (is.null(male_res)) return(NULL)

  tibble(
    code = fit$code,
    name = fit$name,
    region = fit$region,
    onset = onset,
    # Male
    m_median = male_res$median,
    m_lower = male_res$lower,
    m_upper = male_res$upper,
    m_p_pos = male_res$p_positive,
    # Female
    f_median = if (!is.null(female_res)) female_res$median else NA_real_,
    f_lower = if (!is.null(female_res)) female_res$lower else NA_real_,
    f_upper = if (!is.null(female_res)) female_res$upper else NA_real_,
    f_p_pos = if (!is.null(female_res)) female_res$p_positive else NA_real_,
    # OLS comparison
    aic_improvement = fit$aic_improvement
  )
})
```

### Bayesian summary table

```{r}
#| label: tbl-bayesian-summary
#| tbl-cap: "Bayesian impulse-decay estimates for all HMD populations. Posterior median and 95% credible interval for the impulse coefficient (males), posterior probability of positive excess mortality P(γ₀ > 0), and the corresponding female estimate. Ranked by male posterior probability."

bayes_results |>
  arrange(desc(m_p_pos)) |>
  transmute(
    Country = name,
    Region = region,
    Onset = as.character(onset),
    `Male γ₀ [95% CI]` = sprintf("%.3f [%.3f, %.3f]", m_median, m_lower, m_upper),
    `P(γ₀>0) male` = sprintf("%.3f", m_p_pos),
    `Female γ₀ [95% CI]` = ifelse(!is.na(f_median),
      sprintf("%.3f [%.3f, %.3f]", f_median, f_lower, f_upper), "—"),
    `P(γ₀>0) female` = ifelse(!is.na(f_p_pos), sprintf("%.3f", f_p_pos), "—"),
    `AIC impr.` = sprintf("%.0f", aic_improvement)
  ) |>
  knitr::kable()
```

### Forest plot of Bayesian estimates

@fig-bayesian-forest shows the posterior median and 95% credible interval for the impulse coefficient for males (blue) and females (red) across all populations, ranked by male posterior probability.

```{r}
#| label: fig-bayesian-forest
#| fig-cap: "Bayesian impulse-decay coefficient estimates (posterior median and 95% CI) for all HMD populations. Blue = male, red = female. Populations ordered by male posterior probability of positive excess mortality. Vertical dashed line at zero."
#| fig-width: 10
#| fig-height: 14

forest_data <- bayes_results |>
  arrange(m_p_pos) |>
  mutate(country_f = factor(name, levels = name))

male_forest <- forest_data |>
  select(country_f, median = m_median, lower = m_lower, upper = m_upper) |>
  mutate(sex = "Male")

female_forest <- forest_data |>
  filter(!is.na(f_median)) |>
  select(country_f, median = f_median, lower = f_lower, upper = f_upper) |>
  mutate(sex = "Female")

bind_rows(male_forest, female_forest) |>
  ggplot(aes(x = median, y = country_f, colour = sex)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.5),
                  size = 0.3, linewidth = 0.4) +
  scale_colour_manual(values = c("Male" = "steelblue", "Female" = "coral")) +
  labs(x = "Impulse-decay coefficient (posterior median + 95% CI)",
       y = NULL, colour = "Sex") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 8))
```

### Classification by posterior evidence

```{r}
#| label: tbl-bayesian-classification
#| tbl-cap: "Classification of HMD populations by Bayesian posterior evidence. 'Strong positive' = P(γ₀>0) > 0.95 for males; 'Strong negative' = P(γ₀>0) < 0.05; 'Inconclusive' otherwise."

bayes_results |>
  mutate(
    classification = case_when(
      m_p_pos > 0.95 & !is.na(f_p_pos) & f_p_pos < 0.5 ~ "Male-specific disruption",
      m_p_pos > 0.95 & !is.na(f_p_pos) & f_p_pos > 0.95 ~ "Both-sex disruption",
      m_p_pos > 0.95 ~ "General positive disruption",
      m_p_pos < 0.05 ~ "Negative (improving mortality)",
      TRUE ~ "Inconclusive"
    )
  ) |>
  count(classification) |>
  knitr::kable()
```
